---
title: "DL Exploration 02"
author: "David Lovell"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
# Set options to control the behaviour of knitr, dplyr, etc
knitr::opts_chunk$set(echo = FALSE, comment=NA, warning = FALSE, message = FALSE)
options(knitr.table.format = "html", dplyr.summarise.inform = FALSE, tidyverse.quiet = TRUE, width = 100)
```

```{r load-libraries, warning=FALSE, message=FALSE, include=FALSE}
rm(list = ls())
library(tidyverse)
library(scales)    # For trans_format()
library(patchwork)
```

# About this exploration

The aim of this code is to explore _ratios_ of components. Here's a bit of notation to help us along.

* $i$ is the (row) index of each compositional observation
* $j$ and $k$ are the used to index different components (columns)
* $a_{ij}$ is the _actual_ proportion of component $j$ in composition $i$
* $p_{ij}$ is the _predicted_ proportion of component $j$ in composition $i$
  * I'm going to leave out the index of the model making that prediction... already got enough subscripts!
* $r_{ij} = \frac{p_{ij}}{a_{ij}}$ is the _multiplicative residual of component $j$ in composition $i$
  * If $r_{ij} < 1$, we have underestimated the abundance of that component
  * If $r_{ij} = 1$, we have correctly estimated the abundance of that component
  * If $r_{ij} > 1$, we have overestimated the abundance of that component
  
  
I write

* $a_{ij/k} = \frac{a_{ij}}{a_{ik}}$, the ratio of _actual_ components $j$ and $k$ in composition $i$
* $p_{ij/k} = \frac{p_{ij}}{p_{ik}}$, the ratio of _predicted_ components $j$ and $k$ in composition $i$
* $r_{ijk} = \frac{p_{ij/k}}{a_{ij/k}}$, is the multiplicative residual of components $j$ and $k$ in composition $i$

I'm interested to plot how well different models estimate the _ratios_ of components. My plan is to plot $r_{ijk}$ vs $p_{ij/k}$ for each model.

# Read in Khoa's data
This was generated in `DL-Exploration01.Rmd`


```{r}
khoa <- readRDS("./data/khoa.RDS")
#saveRDS(khoa.long, "./data/khoa.long.RDS")
```

# Figure out how to rearrange the data

Let's make life a bit easier to start with and focus on components `A-C` of the first observation of `Model 1`:

```{r}
khoa %>% head(2) %>% select(model, ID, starts_with(LETTERS[1:3])) -> df
```

```{r eval=FALSE, echo=FALSE}
df %>% 
  pivot_longer(cols=contains("predicted"), names_to = "Predicted.component", values_to = "Predicted") %>%
  pivot_longer(cols=contains("actual"),    names_to = "Actual.component",    values_to = "Actual") 
  
```


Stack Overflow to the rescue on [how to make tibble values hold matrices](https://stackoverflow.com/questions/56478279/creating-data-frames-where-one-column-contains-matrices). Note that if we use `as.matrix()` instead of `matrix()` there is different behaviour

```{r}
df %>% rowwise() %>%
mutate(
  actual   =list(matrix(select(., ends_with(".actual")), nrow=1, dimnames=list(NULL, component=LETTERS[1:3]))),
  predicted=list(matrix(select(., ends_with(".predicted"))))
) %>%
  select(model, ID, actual, predicted) -> df.matrix

df.matrix
df.matrix$actual[1][[1]]
```

Let's have a bit more of a play to get the syntax right
```{r}
df %>% rowwise() %>%
mutate(
  actual   =list(matrix(c(A.predicted, B.predicted, C.predicted), nrow=1, dimnames=list(NULL, component=LETTERS[1:3]))),
  predicted1=list(matrix(select(., ends_with(".predicted")))),
  predicted2=list(matrix(c_across(A.predicted:C.predicted))),
  predicted3=list(matrix(c_across(ends_with(".predicted"))))

) %>%
  select(actual, predicted1, predicted2, predicted3) -> df.matrix

df.matrix
df.matrix$actual[1][[1]]
df.matrix$predicted1[1][[1]]
df.matrix$predicted2[1][[1]]
df.matrix$predicted3[1][[1]]
```

OK. Here's what we want
```{r}
df %>% rowwise() %>%
mutate(
  actual    =list(matrix(c_across(ends_with(".actual"   )), nrow=1, dimnames=list(NULL, component=LETTERS[1:3]))),
  predicted =list(matrix(c_across(ends_with(".predicted")), nrow=1, dimnames=list(NULL, component=LETTERS[1:3]))),
  `a_j/k`   =list(outer(actual[1,]   , actual[1,],    FUN="/")),
  `p_j/k`   =list(outer(predicted[1,], predicted[1,], FUN="/")),
  `r_j/k`   =list(`p_j/k` / `a_j/k`)
) %>%
  select(-ends_with(".actual"), -ends_with(".predicted")) -> df.ratios

df.ratios$actual
df.ratios$predicted
df.ratios$`a_j/k`
df.ratios$`p_j/k`
df.ratios$`r_j/k`
```

Promising, but how on earth do we extract the matrix elements back into appropriately named numeric variables?
```{r}
df.ratios %>% select(`r_j/k`) %>% rowwise() %>% unnest_wider(`r_j/k`) %>% colnames
```

Wow! Promising, but look closely and you will see that this tibble has only 3 variables, each of which is a vector.

```{r}
df.ratios %>% select(`r_j/k`) %>% rowwise() %>% unnest_wider(`r_j/k`) %>% colnames
```

What I want is nine columns. Hmm. Let's try a simple example:
```{r}
dimnames <- list(rows=letters[1:3], cols=letters[24:26])
tibble(
  ID = LETTERS[1:4],
  m = list(
    matrix(1:9,   nrow=3, byrow=TRUE, dimnames = dimnames),
    matrix(11:19, nrow=3, byrow=TRUE, dimnames = dimnames),
    matrix(21:29, nrow=3, byrow=TRUE, dimnames = dimnames),
    matrix(31:39, nrow=3, byrow=TRUE, dimnames = dimnames)
  )
) -> m.df
```

```{r}
m.df$m[1]
m.df$m[2]
m.df$m[3]
m.df %>% unnest_wider(m) 
```

hmmm... could try extracting the lower triangle with names?

https://www.tidyverse.org/blog/2020/03/dplyr-1-0-0-summarise/

```{r}
named.lower.tri <- function(m){
  indices <- which(lower.tri(m), arr.ind=TRUE)
  tibble(
    #row=rownames(m)[indices[,1]],
    row=rownames(m)[indices[,1]],
    col=colnames(m)[indices[,2]],
    value=m[indices]
  )
}
```

```{r}
# https://www.tidyverse.org/blog/2020/02/glue-strings-and-tidy-eval/
named.lower.tri <- function(m, prefix="m", values_to="val"){
  indices <- which(lower.tri(m), arr.ind=TRUE)
  tibble(
    "{ prefix }_row"         := rownames(m)[indices[,1]],
    "{ prefix }_col"         := colnames(m)[indices[,2]],
    "{ prefix }_{values_to}" := m[indices]
  )
}
```


```{r}
m.df$m[1][[1]]
m.df$m[1][[1]] %>% named.lower.tri()
```

```{r}
m.df %>% rowwise %>%
  summarise(ID=ID, m=named.lower.tri(m)) %>% unnest(m)
```

```{r}
df.ratios %>% rowwise %>%
  summarise(model=model, ID=ID, named.lower.tri(`a_j/k`, "a", "j/k"), named.lower.tri(`p_j/k`, "p", "j/k"))
```


```{r}
df.ratios %>% rowwise %>%
  summarise(model=model, ID=ID, a=named.lower.tri(`a_j/k`), r=named.lower.tri(`r_j/k`)) # %>% unnest(c(a,r))
```

```{r}
df.ratios %>% group_by(ID) %>%
  summarise(model=model, a=named.lower.tri(`a_j/k`, "a", "j/k"), r=named.lower.tri(`r_j/k`)) 
```


```{r}
df.test <- tibble(
  grp = rep(1:2, each = 5), 
  x = c(rnorm(5, -0.25, 1), rnorm(5, 0, 1.5)),
  y = c(rnorm(5, 0.25, 1), rnorm(5, 0, 0.5)),
)

quibble2 <- function(x, q = c(0.25, 0.5, 0.75)) {
  tibble("{{ x }}" := quantile(x, q), "{{ x }}_q" := q)
}

df.test %>% 
  group_by(grp) %>% 
  summarise(quibble2(y, c(0.25, 0.5, 0.75)))
```

# Let's try a different approach: reshaping pairs of actual and predicted components

Khoa's (cut-down data looks like this:
```{r}
df
```

which we can reshape into
```{r}
df %>% rowwise() %>%
mutate(
  actual    =list(matrix(c_across(ends_with(".actual"   )), nrow=1, dimnames=list(NULL, component=LETTERS[1:3]))),
  predicted =list(matrix(c_across(ends_with(".predicted")), nrow=1, dimnames=list(NULL, component=LETTERS[1:3])))
) %>%
  select(-ends_with(".actual"), -ends_with(".predicted")) -> df.ap

df.ap
df.ap$actual
df.ap$predicted

```

I'd like to try getting this data into the following form:

| model | ID  | j | c_j | a_j      | p_j      | k | c_k | a_k      | p_k     |
|-------|-----|---|-----|----------|----------|---|-----|----------|---------|
| 1     | ID1 | 2 | B   | 0.139373 | 0.05226  | 1 | A   | 0.059128 | 0       |
| 1     | ID1 | 3 | C   | 0.05     | 0.093991 | 1 | A   | 0.059128 | 0       |
| 1     | ID1 | 3 | C   | 0.05     | 0.093991 | 2 | B   | 0.139373 | 0.05226 |

where the rows $j$ and columns $k$ correspond to the elements in the lower triangle.


Let's get the indices of the lower triangle of an $n \times n$ matrix:
```{r}
lower.tri.indices <- function(n){
  indices <- which(lower.tri(matrix(nrow=n, ncol=n)), arr.ind=TRUE)
  tibble(
    j=indices[,1],
    k=indices[,2],
  )
}

lower.tri.indices(3)
```

Now let's try using summarise to print the actual and predicted values in rows:

```{r}
df.ratios %>% rowwise %>%
  summarise(model=model, ID=ID, actual=c(actual), predicted=c(predicted))
```

OK, now let's write a function to get the lower triangle indices for each row of `df.ratios`

```{r}
jk <- lower.tri.indices(3)
df.ratios %>% rowwise %>%
  summarise(
    model=model, ID=ID, jk,
    a_j=c(actual)[j],    a_k=c(actual)[k],
    p_j=c(predicted)[j], p_k=c(predicted)[k]
)
```

Woot! That's it!
