---
title: "DL Exploration 02"
author: "David Lovell"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
# Set options to control the behaviour of knitr, dplyr, etc
knitr::opts_chunk$set(echo = FALSE, comment=NA, warning = FALSE, message = FALSE)
options(knitr.table.format = "html", dplyr.summarise.inform = FALSE, tidyverse.quiet = TRUE, width = 100)
```

```{r load-libraries, warning=FALSE, message=FALSE, include=FALSE}
rm(list = ls())
library(tidyverse)
library(scales)    # For trans_format()
library(viridis)
library(plotly)
```

# About this exploration

The aim of this code is to explore _ratios_ of components. Here's a bit of notation to help us along.

* $i$ is the (row) index of each compositional observation
* $j$ and $k$ are the used to index different components (columns)
* $a_{ij}$ is the _actual_ proportion of component $j$ in composition $i$
* $p_{ij}$ is the _predicted_ proportion of component $j$ in composition $i$
  * I'm going to leave out the index of the model making that prediction... already got enough subscripts!
* $r_{ij} = \frac{p_{ij}}{a_{ij}}$ is the _multiplicative residual of component $j$ in composition $i$
  * If $r_{ij} < 1$, we have underestimated the abundance of that component
  * If $r_{ij} = 1$, we have correctly estimated the abundance of that component
  * If $r_{ij} > 1$, we have overestimated the abundance of that component
  
  
I write

* $a_{ij/k} = \frac{a_{ij}}{a_{ik}}$, the ratio of _actual_ components $j$ and $k$ in composition $i$
* $p_{ij/k} = \frac{p_{ij}}{p_{ik}}$, the ratio of _predicted_ components $j$ and $k$ in composition $i$
* $r_{ijk} = \frac{p_{ij/k}}{a_{ij/k}}$, is the multiplicative residual of components $j$ and $k$ in composition $i$

I'm interested to plot how well different models estimate the _ratios_ of components. My plan is to plot $r_{ijk}$ vs $p_{ij/k}$ for each model.

# Read in Khoa's data
This was generated in `DL-Exploration03.Rmd`

```{r}
khoa.ratios <- readRDS("./data/khoa.ratios.RDS")
```

# Look at all the ratios for models 1 to 8

```{r}
npoints <- 100
xlim <- c(1e-4,1e0)
ylim <- c(1e-2, 1e2)

clip <- function(x, low, high){
  pmin(high, pmax(low, x)) 
}
```

## Plot with points

```{r eval=FALSE, echo=FALSE}
for(Model in 1:8){
  khoa.ratios %>% filter(model==Model, is.finite(`r_j/k`)) %>% # slice_sample(n=npoints) %>% 
    mutate(
      c_j = LETTERS[j],
      c_k = LETTERS[k]
    )  %>% 
    ggplot(
      aes(
        x=clip(`p_j/k`, xlim[1], xlim[2]),
        y=clip(`r_j/k`, ylim[1], ylim[2]))
    ) +
    geom_hline(yintercept=1,  lty=2) + 
    geom_point(alpha=0.25, size=0.5) + 
    #geom_smooth(method = "loess") + 
    scale_x_log10(labels = trans_format("log10", math_format(10^.x))) +
    scale_y_log10(labels = trans_format("log10", math_format(10^.x)), breaks=10^seq(from=-4, to=4, by=2)) +
    coord_equal(xlim=xlim, ylim=ylim) + 
    theme(axis.text.x = element_text(angle = 90)) +
    facet_grid(c_j ~ c_k, labeller=label_both) + 
    labs(
      x="p_j/k", y="r_j/k",
      title=sprintf("Model %d: r_j/k vs p_j/k", Model),
      subtitle=sprintf("Showing points on a log-log scale from x=%0.1e to x=%0.1e", xlim[1], xlim[2]),
      caption="Points with infinite residuals have been removed. Values outside the range of an axes have been set to the nearest in range value."
    ) -> p
  
  ggsave(sprintf("plots/Model%d-multiplicative-residuals.png", Model), plot=p, width = 10, height=10, dpi=600)
}
```

## Plot with bins

Binwidths and color scales can help reveal details
```{r}
for(Model in 1:8){
  khoa.ratios %>% filter(model==Model, is.finite(`r_j/k`)) %>%  # slice_sample(n=npoints) %>% 
    mutate(
      c_j = LETTERS[j],
      c_k = LETTERS[k]
    )  %>% 
    ggplot(
      aes(
        x=clip(`p_j/k`, xlim[1], xlim[2]),
        y=clip(`r_j/k`, ylim[1], ylim[2]))
    ) +
    geom_hline(yintercept=1,  lty=2) + 
    geom_bin2d(bins=150)+
    scale_fill_viridis(trans = "log10", option='H') +
    scale_x_log10(labels = trans_format("log10", math_format(10^.x))) +
    scale_y_log10(labels = trans_format("log10", math_format(10^.x)), breaks=10^seq(from=-4, to=4, by=2)) +
    coord_equal(xlim=xlim, ylim=ylim) + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
    facet_grid(c_j ~ c_k) + 
    labs(
      title=sprintf("Model %d", Model),
      subtitle=expression(paste(r[j/k], " vs ", p[j/k])),
      x=expression(p[j/k]), y=expression(r[j/k]),
      caption="Points with infinite residuals have been removed. Values outside the range of an axes have been set to the nearest in range value."
    ) -> p

  ggsave(sprintf("plots/Model%d-multiplicative-residuals-.png", Model), plot=p, width = 10, height=10, dpi=600)
}
```

```{r eval=FALSE, echo=FALSE}
ggplot(tibble(x=1,y=1), aes(x=x, y=y)) + 
  labs(
    title=expression(paste("Model ", .(Model), ": ", r[j/k], " vs ", p[j/k])),
    x=expression(p[j/k]), y=expression(r[j/k])
  )
```

# Zeros!



```{r}
head(khoa.ratios,1000) %>% select(model, a_j, a_k, p_j, p_k) %>%
  mutate(
    bits=
      8*(a_j != 0) +
      4*(a_k != 0) +
      2*(p_j != 0) +
      1*(p_k != 0)
  ) %>%
  group_by(model) %>%
  count(bits)
```

```{r}
khoa.ratios %>% select(model, a_j, a_k, p_j, p_k) %>%
  mutate(
    bitstring=sprintf(
      "%s%s%s%s",
      ifelse(a_j != 0,  "1", "0"),
      ifelse(a_k != 0,  "1", "0"),
      ifelse(p_j != 0,  "1", "0"),
      ifelse(p_k != 0,  "1", "0")
    ) 
  ) %>%
  group_by(model) %>%
  count(bitstring) -> khoa.ratios.zeros
```


```{r}
khoa.ratios.zeros %>% ungroup %>%
  mutate(
    bitstring = fct_reorder(factor(bitstring), n)
  ) %>%
  ggplot(aes(y=n, x=bitstring, color=model, group=model)) + geom_point() + geom_line() + 
  scale_y_log10() + coord_flip()
```


```{r}
khoa <- readRDS("./data/khoa.RDS")
```

```{r}
khoa %>% head(0) %>% dplyr::select(ends_with(".actual"   )) %>% ncol -> n
start.time <- Sys.time()
khoa %>%  # head(1000) %>%
  mutate(
    ID=factor(ID)
  ) %>%
  rowwise %>%
  mutate(
    actual    =list(c(A.actual,B.actual,C.actual,D.actual,E.actual,F.actual,G.actual,H.actual,I.actual)),
    predicted =list(c(A.predicted,B.predicted,C.predicted,D.predicted,E.predicted,F.predicted,G.predicted,H.predicted,I.predicted))
  )%>%
  dplyr::select(model, ID, actual, predicted)  %>%
  summarise(
    model=model, j=1:n, actual=c(actual), predicted=c(predicted)
  ) %>%
  mutate(
    zero=factor(
      case_when(
        (actual != 0) & (predicted != 0)  ~ "neither",
        (actual == 0)                     ~ "actual",
                        (predicted == 0)  ~ "predicted",
        TRUE                              ~ "both"
      )
    )
  ) %>%
  group_by(model) %>%
  count(zero)  -> khoa.zeros

end.time <- Sys.time()
time.taken <- round(end.time - start.time,2)
time.taken
```

```{r}
khoa.zeros %>% 
  mutate(
    zero=fct_reorder(zero, n)
  ) %>%
ggplot(aes(y=n, x=zero, color=model, group=model)) + geom_point() + geom_line() + 
  scale_y_log10() + 
  labs(title="Numbers of zero actual or predicted components") +
  coord_flip() -> p
ggplotly(p)
```

