---
title: "David's initial exploration"
author: "David Lovell"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
# Set options to control the behaviour of knitr, dplyr, etc
knitr::opts_chunk$set(echo = FALSE, comment=NA, warning = FALSE, message = FALSE)
options(knitr.table.format = "html", dplyr.summarise.inform = FALSE, tidyverse.quiet = TRUE, width = 100)
```

```{r load-libraries, warning=FALSE, message=FALSE, include=FALSE}
rm(list = ls())
library(tidyverse)
```

# Try reading in one of Khoa's files

Khoa has sent us a range of `.csv` files containing the predictions of the proportion of different cell types in a sample.
He's sent us the predictions made by 8 different models, and he's sent us the ground truth.

Our first job is to ensure we can ingest and interpret these data files.

Let's try reading `method_1.csv`:

```{r}
read_tsv("M1.csv")
```

Cool! That works. It is actually a TAB separated value file, not a `.csv` and the first column has no name (the string `...1` is used as a legal placeholder) so we can fix that with...

```{r}
read_tsv("M1.csv") %>% rename(ID="...1")
```

# Try reading in the truth

So, let's see what we have here:
```{r}
read_tsv("truth.csv")
```

OK! So we have a similar file of IDs and (I hope) proportions.

# Combine model predictions with ground truth

I think a good way to do this is to do a database join using the ID as a key... I'm not quite sure how to do this, so I will have a play with some cut down data:
```{r}
read_tsv("M1.csv", n_max=10, show_col_types = FALSE) %>% rename(ID="...1") %>% select(ID, A, B) -> method1
read_tsv("truth.csv",    n_max=10, show_col_types = FALSE) %>% rename(ID="...1") %>% select(ID, A, B) -> truth
```

Let's do a _full_ join (I think!) using the suffixes `.pred` for "predicted" and `.act` for "actual"

```{r}
full_join(method1, truth, by="ID", suffix=c(".pred", ".act"))
```

Looking good! Now let's see how we might ingest all 8 models' worth of predictions. If the data is small enough, my usual strategy is to create a single data frame that I can then subset. If the data is too big for that, I'm forced to work in chunks.

# Let's try reading in all the model predictions

There is probably a neater more compact way to do this, but I'm happy to be verbose.

```{r}
n_max <- 10
bind_rows(
  .id="model",
  read_tsv("M1.csv", n_max=n_max, show_col_types = F),
  read_tsv("M2.csv", n_max=n_max, show_col_types = F)
) %>% 
  rename(ID="...1") -> predictions

```

Now get the truth:
```{r}
read_tsv("truth.csv",    n_max=n_max, show_col_types = F) %>% rename(ID="...1")  -> truth
```

Now join to the truth:
```{r}
full_join(predictions, truth, by="ID", suffix=c(".pred", ".act")) -> khoa
```

# Show how to convert to long format

`ggplot()` requires data in _long_ format, i.e., one measurement per row. What I want to get to is data in the form of

```
model ID         component predicted actual
1     96b0a24b…  A         0         0.0591
...
```

This is a job for `pivot_longer()`. I'm going to approach this in two stages maybe...
```{r}
pivot_longer(
  khoa, 
  cols= -c(model, ID), 
  names_to = c(".value", "label"), 
  names_pattern = "(.).(.*)")
```

Close but no cigar!

```{r}
pivot_longer(
  khoa, 
  cols= -c(model, ID), 
  names_to = c(".value", "label"), 
  names_pattern = "(.).(.*)")
```



```{r}
pnl <- tibble(
  x = 1:4,
  a = c(1, 1,0, 0),
  b = c(0, 1, 1, 1),
  y1 = rnorm(4),
  y2 = rnorm(4),
  z1 = rep(3, 4),
  z2 = rep(-2, 4),
)
```

```{r}
pnl %>% 
  pivot_longer(
    !c(x, a, b), 
    names_to = c(".value", "time"), 
    names_pattern = "(.)(.)"
  )
```




```{r}
tibble(
  x = 1:4,
  a = c(1, 1,0, 0),
  b = c(0, 1, 1, 1),
  y.pred = rnorm(4),
  y.act = rnorm(4),
  z1 = rep(3, 4),
  z2 = rep(-2, 4),
) %>% 
  pivot_longer(
    !c(x, a, b), 
    names_to = c(".value", "time"), 
    names_pattern = "(.).(.*)"
  )

```



```{r}
khoa %>% 
  pivot_longer(
    cols= -c(model, ID), 
    names_to = c(".value", "label"), 
    names_pattern = "(.).(.*)") %>%
  pivot_longer(cols=-c(model, ID, label), names_to = "component") %>%
  pivot_wider(names_from = label) -> khoa.long
```


```{r}
ggplot(khoa.long, aes(x=act, y=pred, group=fct_cross(model, ID))) +
  geom_point() + geom_line() + coord_equal(xlim=c(0,1), ylim=c(0,1)) + facet_grid(model ~ component)
```



```{r}
ggplot(khoa.long, aes(x=act, y=pred)) +
  geom_abline(slope=1, color="white") + 
  geom_point() + 
  coord_equal(xlim=c(0,1), ylim=c(0,1)) + 
  facet_grid(model ~ component)
```







