---
title: "DL Exploration 02"
author: "David Lovell"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
# Set options to control the behaviour of knitr, dplyr, etc
knitr::opts_chunk$set(echo = FALSE, comment=NA, warning = FALSE, message = FALSE)
options(knitr.table.format = "html", dplyr.summarise.inform = FALSE, tidyverse.quiet = TRUE, width = 100)
```

```{r load-libraries, warning=FALSE, message=FALSE, include=FALSE}
rm(list = ls())
library(tidyverse)
library(scales)    # For trans_format()
library(patchwork)
```

# About this exploration

The aim of this code is to explore _ratios_ of components. Here's a bit of notation to help us along.

* $i$ is the (row) index of each compositional observation
* $j$ and $k$ are the used to index different components (columns)
* $a_{ij}$ is the _actual_ proportion of component $j$ in composition $i$
* $p_{ij}$ is the _predicted_ proportion of component $j$ in composition $i$
  * I'm going to leave out the index of the model making that prediction... already got enough subscripts!
* $r_{ij} = \frac{p_{ij}}{a_{ij}}$ is the _multiplicative residual of component $j$ in composition $i$
  * If $r_{ij} < 1$, we have underestimated the abundance of that component
  * If $r_{ij} = 1$, we have correctly estimated the abundance of that component
  * If $r_{ij} > 1$, we have overestimated the abundance of that component
  
  
I write

* $a_{ij/k} = \frac{a_{ij}}{a_{ik}}$, the ratio of _actual_ components $j$ and $k$ in composition $i$
* $p_{ij/k} = \frac{p_{ij}}{p_{ik}}$, the ratio of _predicted_ components $j$ and $k$ in composition $i$
* $r_{ijk} = \frac{p_{ij/k}}{a_{ij/k}}$, is the multiplicative residual of components $j$ and $k$ in composition $i$

I'm interested to plot how well different models estimate the _ratios_ of components. My plan is to plot $r_{ijk}$ vs $p_{ij/k}$ for each model.

# Read in Khoa's data
This was generated in `DL-Exploration01.Rmd`

```{r}
khoa <- readRDS("./data/khoa.RDS")
```

# Rearrange the data

## Baby steps

This was figured out in `DL-Exploration02.Rmd`
Let's focus on components `A-C` of the first observation of `Model 1`:

```{r eval=FALSE, echo=FALSE}
# Attempt 1
khoa %>% head(2) %>% select(model, ID, starts_with(LETTERS[1:3])) %>%  rowwise %>%
  mutate(
    actual   =list(c(select(., ends_with(".actual")))),
    predicted=list(c(select(., ends_with(".predicted"))))
  ) %>%
  select(model, ID, actual, predicted) -> df
df
df$actual[1][[1]]

# Attempt 2
khoa %>% head(2) %>% select(model, ID, starts_with(LETTERS[1:3])) %>% rowwise %>%
mutate(
  actual   =list(matrix(select(., ends_with(".actual")), nrow=1, dimnames=list(NULL, component=LETTERS[1:3]))),
  predicted=list(matrix(select(., ends_with(".predicted"))))
) %>%
  select(model, ID, actual, predicted) -> df
df
df$actual[1][[1]]

# Attempt 3
khoa %>% head(2) %>% select(model, ID, starts_with(LETTERS[1:3])) %>% rowwise %>%
mutate(
  actual    =list(matrix(c_across(ends_with(".actual"   )), nrow=1, dimnames=list(NULL, component=LETTERS[1:3]))),
  predicted =list(matrix(c_across(ends_with(".predicted")), nrow=1, dimnames=list(NULL, component=LETTERS[1:3])))
)%>%
  select(model, ID, actual, predicted) -> df
df
df$actual[1][[1]]
```


```{r}
khoa %>% head(2) %>% select(model, ID, starts_with(LETTERS[1:3])) %>% rowwise %>%
mutate(
  actual    =list(c_across(ends_with(".actual"   ))),
  predicted =list(c_across(ends_with(".predicted")))
)%>%
  select(model, ID, actual, predicted) -> df

df
```

Here's a function to generate a tibble of lower triangle indices
```{r}
lower.tri.indices <- function(n){
  indices <- which(lower.tri(matrix(nrow=n, ncol=n)), arr.ind=TRUE)
  tibble(
    j=indices[,1],
    k=indices[,2],
  )
}
```

Now let's use that to rearrange our baby data frame

```{r}
jk <- lower.tri.indices(length(df$actual[1][[1]]))
df %>% rowwise %>%
  summarise(
    model=model, ID=ID, jk,
    a_j=c(actual)[j],    a_k=c(actual)[k],
    p_j=c(predicted)[j], p_k=c(predicted)[k]
)
```

## Toddler steps

Now let's use all 9 components of 2 observations
```{r}
khoa %>% head(2) %>% rowwise %>%
mutate(
  actual    =list(c_across(ends_with(".actual"   ))),
  predicted =list(c_across(ends_with(".predicted")))
)%>%
  select(model, ID, actual, predicted) -> df

df
```

```{r}
jk <- lower.tri.indices(length(df$actual[1][[1]]))
df %>% rowwise %>%
  summarise(
    model=model, ID=ID, jk,
    a_j=c(actual)[j],    a_k=c(actual)[k],
    p_j=c(predicted)[j], p_k=c(predicted)[k]
) -> df.rearranged

df.rearranged
```

So the 

* `r nrow(df)` by `r ncol(khoa)` source data frame is rearranged into 
* `r nrow(df.rearranged)` by `r ncol(df.rearranged)` target data frame

that's a `r nrow(df.rearranged) * ncol(df.rearranged) / (nrow(df) * ncol(khoa))` increase...

That's the price we pay for working in ratios!

## Big boy steps

Now let's use all 9 components of 2 observations and calculate some ratios!
```{r}
# Generate the lower triangle indices for the data
khoa %>% head(0) %>% select(ends_with(".actual"   )) %>% ncol %>% lower.tri.indices -> jk

khoa %>% head(2) %>% rowwise %>%
  mutate(
    actual    =list(c_across(ends_with(".actual"   ))),
    predicted =list(c_across(ends_with(".predicted")))
  )%>%
  select(model, ID, actual, predicted) %>%
  summarise(
    model=model, ID=ID, jk,
    a_j=c(actual)[j],    a_k=c(actual)[k],    `a_j/k`=a_j/a_k,
    p_j=c(predicted)[j], p_k=c(predicted)[k], `p_j/k`=p_j/p_k,
                                              `r_j/k`=`p_j/k` / `a_j/k`
  )
```

## Teenager steps

Now let's do the whole dataset and see how long it takes...
```{r}

khoa %>% head(0) %>% select(ends_with(".actual"   )) %>% names %>% cat(sep=",")
khoa %>% head(0) %>% select(ends_with(".predicted")) %>% names %>% cat(sep=",")
# Generate the lower triangle indices for the data
khoa %>% head(0) %>% select(ends_with(".actual"   )) %>% ncol %>% lower.tri.indices -> jk

# This runs slow as a wet week. Better to specify the variables explicitly
# khoa %>% head(2000) %>%  rowwise %>%
#   mutate(
#     actual    =list(c_across(ends_with(".actual"   ))),
#     predicted =list(c_across(ends_with(".predicted")))
#   )%>%
#   select(model, ID, actual, predicted) -> df
# end.time <- Sys.time()
# time.taken <- round(end.time - start.time,2)
# time.taken

start.time <- Sys.time()
khoa %>% 
  mutate(
    ID=factor(ID)
  ) %>%
  rowwise %>%
  mutate(
    actual    =list(c(A.actual,B.actual,C.actual,D.actual,E.actual,F.actual,G.actual,H.actual,I.actual)),
    predicted =list(c(A.predicted,B.predicted,C.predicted,D.predicted,E.predicted,F.predicted,G.predicted,H.predicted,I.predicted))
  )%>%
  select(model, ID, actual, predicted)  %>%
  summarise(
    model=model, ID=ID, jk,
    a_j=c(actual)[j],    a_k=c(actual)[k],    `a_j/k`=a_j/a_k,
    p_j=c(predicted)[j], p_k=c(predicted)[k], `p_j/k`=p_j/p_k,
                                              `r_j/k`=`p_j/k` / `a_j/k`
  ) -> khoa.ratios

end.time <- Sys.time()
time.taken <- round(end.time - start.time,2)
time.taken
```

```{r}
saveRDS(khoa.ratios, "./data/khoa.ratios.RDS")
```

