---
title: "DL Exploration 02"
author: "David Lovell"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
# Set options to control the behaviour of knitr, dplyr, etc
knitr::opts_chunk$set(echo = FALSE, comment=NA, warning = FALSE, message = FALSE)
options(knitr.table.format = "html", dplyr.summarise.inform = FALSE, tidyverse.quiet = TRUE, width = 100)
```

```{r load-libraries, warning=FALSE, message=FALSE, include=FALSE}
rm(list = ls())
library(tidyverse)
library(scales)    # For trans_format()
library(patchwork)
library(viridis)
```

# About this exploration
This is to try out the

* [Color points by density with ggplot2](https://slowkow.com/notes/ggplot2-color-by-density/) and
* [Binning strategies for big data density plots](https://twitter.com/ZanotelliVRT/status/1248292074471948289)

concepts to help get a better sense of some heavily overplotted data

# Read in Khoa's data
This was generated in `DL-Exploration01.Rmd`


```{r}
khoa.long <- readRDS("./data/khoa.long.RDS")
```

# My original attempt
```{r}
npoints <- 10000
xlim <- c(1e-9,1e0)
ylim <- c(1e-4, 1e4)

clip <- function(x, low, high){
  pmin(high, pmax(low, x)) 
}

start.time <- Sys.time()

khoa.long %>% filter(actual > 0, predicted > 0) %>% slice_sample(n=npoints) %>% 
ggplot(
  aes(
    x=clip(predicted, xlim[1], xlim[2]),
    y=clip(predicted/actual, ylim[1], ylim[2]))
  ) +
  geom_hline(yintercept=1,  lty=2) + 
  geom_point(alpha=0.25, size=0.5) + 
  # geom_smooth(method = "loess") + 
  scale_x_log10(labels = trans_format("log10", math_format(10^.x))) +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x)), breaks=10^seq(from=-4, to=4, by=2)) +
  coord_equal(xlim=xlim, ylim=ylim) + 
  theme(axis.text.x = element_text(angle = 90)) +
  facet_grid(model ~ component, labeller=label_both) + 
  labs(
    x="predicted", y="predicted/actual",
    title="Predicted/actual component proportions vs predicted proportions",
    subtitle=sprintf("Showing a sample of %d points on a log-log scale from x=%0.1e to x=%0.1e", npoints, xlim[1], xlim[2]),
    caption="Points with zero actual or predicted values have been removed. Values outside the range of an axes have been set to the nearest in range value."
  )

end.time <- Sys.time()
time.taken <- round(end.time - start.time,2)
time.taken
```

# Using a 2d binning strategy

| npoints | time.taken |
|---------|------------|
| 1e4     | 14s        |
| 1e5     | 15s        |
| 1e6     | 18s        |
| 1.9e6   | 19s        |

```{r}
xlim <- c(1e-9,1e0)
ylim <- c(1e-4, 1e4)
#npoints <- 1e7
khoa.long %>% filter(actual > 0, predicted > 0) %>% nrow -> npoints

start.time <- Sys.time()

khoa.long %>% filter(actual > 0, predicted > 0) %>% # slice_sample(n=npoints) %>% 
ggplot(aes(x=predicted,y=predicted/actual)) +
  geom_hline(yintercept=1,  lty=2) + 
  geom_bin2d(bins=300)+
  scale_fill_viridis(trans = "log10", option='inferno') +
  scale_x_log10(oob=squish, labels = trans_format("log10", math_format(10^.x))) +
  scale_y_log10(oob=squish, labels = trans_format("log10", math_format(10^.x)), breaks=10^seq(from=-4, to=4, by=2)) +
  coord_equal(xlim=xlim, ylim=ylim) + 
  theme(axis.text.x = element_text(angle = 90)) +
  facet_grid(model ~ component, labeller=label_both) + 
  labs(
    x="predicted", y="predicted/actual",
    title="Predicted/actual component proportions vs predicted proportions",
    subtitle=sprintf("Showing a sample of %d points on a log-log scale from x=%0.1e to x=%0.1e", npoints, xlim[1], xlim[2]),
    caption="Points with zero actual or predicted values have been removed. Out of bound values have been squished."
  )

end.time <- Sys.time()
time.taken <- round(end.time - start.time,2)
time.taken

ggsave("plots/relative-vs-actual-binned.png", width = 10, height=10, dpi=600)

```

Alas, the squishing doesn't seem to work so well with the binning. Back to the manual approach!

```{r}
xlim <- c(1e-9,1e0)
ylim <- c(1e-4, 1e4)
#npoints <- 1e7
khoa.long %>% filter(actual > 0, predicted > 0) %>% nrow -> npoints
clip <- function(x, low, high){
  pmin(high, pmax(low, x)) 
}

start.time <- Sys.time()

khoa.long %>% filter(actual > 0, predicted > 0) %>% # slice_sample(n=npoints) %>% 
ggplot(
    aes(
    x=clip(predicted, xlim[1], xlim[2]),
    y=clip(predicted/actual, ylim[1], ylim[2]))
  ) +
  geom_hline(yintercept=1,  lty=2) + 
  geom_bin2d(bins=300)+
  scale_fill_viridis(trans = "log10", option='H') +
  scale_x_log10(oob=squish, labels = trans_format("log10", math_format(10^.x))) +
  scale_y_log10(oob=squish, labels = trans_format("log10", math_format(10^.x)), breaks=10^seq(from=-4, to=4, by=2)) +
  coord_equal(xlim=xlim, ylim=ylim) + 
  theme(axis.text.x = element_text(angle = 90)) +
  facet_grid(model ~ component, labeller=label_both) + 
  labs(
    x="predicted", y="predicted/actual",
    title="Predicted/actual component proportions vs predicted proportions",
    subtitle=sprintf("Showing a sample of %d points on a log-log scale from x=%0.1e to x=%0.1e", npoints, xlim[1], xlim[2]),
    caption="Points with zero actual or predicted values have been removed. Out of bound values have been squished."
  )

end.time <- Sys.time()
time.taken <- round(end.time - start.time,2)
time.taken

ggsave("plots/relative-vs-actual-binned-300-clip.png", width = 10, height=10, dpi=600)

```


```{r}
xlim <- c(1e-9,1e0)
ylim <- c(1e-4, 1e4)
#npoints <- 1e7
khoa.long %>% filter(actual > 0, predicted > 0) %>% nrow -> npoints
clip <- function(x, low, high){
  pmin(high, pmax(low, x)) 
}

start.time <- Sys.time()

khoa.long %>% filter(actual > 0, predicted > 0) %>% # slice_sample(n=npoints) %>% 
ggplot(
    aes(
    x=clip(predicted, xlim[1], xlim[2]),
    y=clip(predicted/actual, ylim[1], ylim[2]))
  ) +
  geom_hline(yintercept=1,  lty=2) + 
  geom_bin2d(bins=150)+
  scale_fill_viridis(trans = "log10", option='H') +
  scale_x_log10(oob=squish, labels = trans_format("log10", math_format(10^.x))) +
  scale_y_log10(oob=squish, labels = trans_format("log10", math_format(10^.x)), breaks=10^seq(from=-4, to=4, by=2)) +
  coord_equal(xlim=xlim, ylim=ylim) + 
  theme(axis.text.x = element_text(angle = 90)) +
  facet_grid(model ~ component, labeller=label_both) + 
  labs(
    x="predicted", y="predicted/actual",
    title="Predicted/actual component proportions vs predicted proportions",
    subtitle=sprintf("Showing a sample of %d points on a log-log scale from x=%0.1e to x=%0.1e", npoints, xlim[1], xlim[2]),
    caption="Points with zero actual or predicted values have been removed. Out of bound values have been squished."
  )

end.time <- Sys.time()
time.taken <- round(end.time - start.time,2)
time.taken

ggsave("plots/relative-vs-actual-binned-150-clip.png", width = 10, height=10, dpi=600)

```



# Additive residuals

```{r}
start.time <- Sys.time()

khoa.long %>% 
ggplot(
    aes(
    x=predicted,
    y=predicted-actual)
  ) +
  geom_hline(yintercept=0,  lty=2) + 
  geom_bin2d(bins=300)+
  scale_fill_viridis(trans = "log10", option='H') +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_grid(model ~ component, labeller=label_both) + 
  labs(
    x="predicted", y="predicted - actual",
    title="(predicted - actual) component proportions vs predicted proportions",
    subtitle="300 bins"
  )

end.time <- Sys.time()
time.taken <- round(end.time - start.time,2)
time.taken

ggsave("plots/residual-vs-predicted-binned-300.png", width = 10, height=10, dpi=600)
```

```{r}
start.time <- Sys.time()

khoa.long %>% 
ggplot(
    aes(
    x=predicted,
    y=predicted-actual)
  ) +
  geom_hline(yintercept=0,  lty=2) + 
  geom_bin2d(bins=150)+
  scale_fill_viridis(trans = "log10", option='H') +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_grid(model ~ component, labeller=label_both) + 
  labs(
    x="predicted", y="predicted - actual",
    title="(predicted - actual) component proportions vs predicted proportions",
    subtitle="150 bins"
  )

end.time <- Sys.time()
time.taken <- round(end.time - start.time,2)
time.taken

ggsave("plots/residual-vs-predicted-binned-150.png", width = 10, height=10, dpi=600)
```

```{r}
khoa.long %>% select(model, predicted, actual) %>% pivot_longer(-model) -> tmp
bind_rows(
  filter(tmp, name=="predicted") %>% select(-name),
  filter(tmp, name!="predicted") %>% select(-name) %>% mutate(model="actual")
) %>%
  mutate(
    value=ifelse(value==0, 1e-16, value)
  ) %>% 
  #filter(is.infinite(value))
  ggplot(aes(x=value)) +
  geom_histogram(breaks=c(1e-17, 10^(-16:0)), closed="right", fill="white", color="black") +
  scale_x_log10() + scale_y_log10() +
  facet_grid(model ~ .) +
  labs(title="Numbers of model predicted and actual values in different ranges")
```

