---
title: "NGP_MethodAnalysis01"
author: "Nicholas Gecks-Preston"
date: "2023-01-31"
output: html_document
---

## Setup and Preabmle
First I will make sure each time the code runs the enviroment and packages are clear.

```{r setup}
rm(list = ls())
if (length((.packages())) > 7){
invisible(lapply(paste0("package:", names(sessionInfo()$otherPkgs)),
                 detach,
                 character.only = TRUE, unload = TRUE))}
```

Next I will Library all the required packages.

```{r, warning=FALSE, results='hide', message=FALSE}
library("tidyverse")
library("plotly")
library("broom")
library("readr")
```

Next I need to read in the files Khoa has given me, I found out from my previous exploration with david that they are in fact not csvs' files but tsvs'

I will do this all at the same time as I have already done the singular case with David

```{r,warning=FALSE, results='hide', message=FALSE}
n_max <- 10
list.files(pattern = "*.csv") %>% lapply(read_tsv, n_max=n_max, show_col_types = FALSE) -> Raw_Data
names(Raw_Data) <- gsub(".csv", "",
                 list.files(pattern = "*.csv",full.names = FALSE),
                 fixed = TRUE)

bind_rows(.id="model", Raw_Data[]) %>% rename(ID="...1") -> Bound_Data


truth <- Bound_Data[which(Bound_Data$model == "truth"),]
predictions <-Bound_Data[which(Bound_Data$model != "truth"),]
rm("Raw_Data","Bound_Data")

full_join(predictions, truth, by="ID", suffix=c(".pred", ".act")) %>% 
  rename(model="model.pred") %>% 
  subset(select = -c(model.act)) %>%
  pivot_longer(
    cols= -c(model, ID), 
    names_to = c(".value", "label"), 
    names_pattern = "(.).(.*)") %>%
  pivot_longer(cols=-c(model, ID, label), names_to = "component") %>%
  pivot_wider(names_from = label) -> Khoa

rm("truth","predictions")

```

Oki so now we can easily plot this with gg plot I will borrow some of David's Code

```{r}
ggplot(Khoa, aes(x=act, y=pred)) +
  geom_abline(slope=1, color="Red") + 
  geom_point() + 
  coord_equal(xlim=c(0,1), ylim=c(0,1)) + 
  facet_grid(model ~ component) +
  theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank())+
  xlab("Ground Truth")+
  ylab("Model Prediction")
```

Now we have a graph let's quantify the accuracy of the predictions.
